Computing MEM seeding probabilities

### 2019-02-13 MEM.tex
Added section about practical considerations.

The focus of this document is not to solve the true location problem
\textit{per se}, but rather to evaluate whether a solution is correct. At
first sight, whether we can map reads to their true location seems to
depend only on their lengths and on the error rate of the instrument.
However, there is another important factor to consider: most genomes
contain repeated sequences, \textit{i.e.} relatively large subsequences
(from $\sim20$~bp to more than 10~kb) that are present at more than one
location. So we cannot map the fragment with certainty if there exists an
exact copy of the sequence at somewhere else in the genome, because we
cannot know which of the two copies was sequenced.

@@

Because sequencing errors can convert a DNA fragment into one of its
duplicates, the \emph{true} location of a DNA fragment is not always the
\emph{best} (as measured by the identity between the sequence and the
candidate location). This leads us to consider the following question: how
can we identify the \emph{best} location of the fragment in the genome?

This is the \emph{best location problem}. It ammounts to finding the
optimal alignment between two sequences, and for this reason has received
substantial attention in bioinformatics. It is relevant for the present
discussion because the candidate solution of the \emph{true} location
problem is always the \emph{best} location. The next section will further
clarify the relationship between the two problems.

@@

There exists exact algorithms to solve the best location
problem\cite{pmid7265238,pmid5420325}, but they are too slow to process
the large amount of data generated by modern sequencers. Instead, one uses
heuristic methods, \textit{i.e.} algorithms that run fast, but may not
return the best location\cite{Waterman1984}. Heuristic mapping algorithms
thus have three possible outcomes: $i.$ the DNA fragment is mapped to the
best location, $ii.$ the DNA fragment is mapped to a suboptimal location,
$iii.$ the DNA fragment is not mapped at all.

@@

If seeds can be discovered fast, we can quickly zoom in on a small set of
candidate locations, and use slower exact algorithms on a reduced data
set. The disadvantage is that the target location is not always in the
candidate set. In this case, it cannot be discovered and the outcome is
either a false positive or a false negative.

This imposes a trade-off between speed and accuracy between each candidate
location must be tested with a slow algorithm. If the seeding step is set
to produce many candidates, the target is likely to be among them and it
will be discovered. The downside is that this will take long. Conversely,
if the seeding step is set to produce few candidates, the process will
terminate faster but the target is less likely to sit among the candidates
so it is less likely to be discovered.

@@

This framework was developed for the best location problem, but it also
holds for the true location problem. Actually, the process is exactly the
same but the outcomes $i.$ and $ii.$ are different. The first, called a
``true positive'' or an ``on-target hit'', is that the DNA fragment is
mapped to the \emph{true} location; the second, called a ``false
positive'' or an ``off-target hit'' is that the DNA fragment is mapped
somewhere else in the genome. A true positive thus requires that the
target (the true location) is in the candidate set, and that it is the
optimum.

@@

In general, seeds do not have to be exact matches\cite{pmid20460430}. For
instance, the algorithm PatternHunter\cite{pmid11934743} uses ``spaced
seeds'' that tolerate mismatches\footnote{The method is actually covered
by a patent (reference US20070088510A1).}. However, in this document, we
will always assume that a seed is an exact match with no gap. In other
words, a seed is a perfect match between a part of the read and a genomic
site.

@@

Before going further, we need to move a practical consideration out of the
way. It stands to reason that very short matches between the read and the
genome are uninformative. For instance, $99.7\%$ of all the possible
12-mers are present in the reference human genome, so an exact match of
size 12 or lower says nothing about the location of the DNA fragment.
In what follows, we will assume that subsequences of the read can only
match the original DNA sequence and/or its duplicates, but not the rest of
the genome. This will simplify the discussion without loss of generality
because non homologous matches are short and they can be ignored.

@@

@@

The last assumption is that MEM seeds can match only the target or one of
its duplicates. This is not the case: MEM seeds overlapping a sequencing
error often match random locations of the genome. When this occurs, the
MEM seeds are typically small and the alignment score at the candidate
location is very bad because the sequences are unrelated. So in practice,
it is relatively easy to detect this case and reject the location, even if
it is the best in the candidate set.

The proposed methods are compliant with the demand for speed of the
mapping algorithms. One needs to compute the probabilities only once for a
given value of $N$ and $\mu$ (the error rate $p$ is known and constant).
For $N > 20$, the iterative method is usually prohibitive and we need to
use Monte Carlo sampling instead, which for reads of size $< 200$
nucleotides keeps the computations around 1 second on modern hardware. The
values of $N$ and $\mu$ can be binned in intervals of interest so that
there are only $\approx 100$ pairs for a total cost of a few minutes per
run. Considering that mappers can process at most $10,000$ reads per
second, the computer time of mapping a sequencing run of 250 million reads
is over 7 hours, two orders of magnitude larger than the time required to
estimate the probabilities of error.

>>>>

The focus of this document is not to solve the true location problem
\textit{per se}, but rather to evaluate whether a given solution is
correct. At first sight, whether we can map a read to its true location
seems to depend only on its length and on the error rate of the
instrument. However, there is another important factor to consider: most
genomes contain repeated sequences, \textit{i.e.} relatively large
subsequences that are present at more than one location. So we cannot map
the fragment with certainty if there exists an exact copy of the sequence
somewhere else in the genome, because we cannot know which of the two
copies was sequenced.

@@

Because sequencing errors can convert a DNA fragment into one of its
duplicates, the \emph{true} location of a DNA fragment is not always the
\emph{best} (as measured by the identity between the sequence and the
candidate location). This leads to the following question: how can we
identify the \emph{best} location of the fragment in the genome?

We will refer to this as the \emph{best location problem}. It amounts to
finding the optimal alignment between two sequences, and for this reason
has received substantial attention in bioinformatics. The distinction is
relevant for the present discussion because the candidate solution of the
\emph{true} location problem can only be the \emph{best} location. So in
order to solve the true location problem, we must first solve the best
location problem.

@@

There exists exact algorithms to solve the best location
problem\cite{pmid7265238,pmid5420325}, but they are too slow to process
the large amount of data generated by modern sequencers. Instead, one uses
heuristic methods, \textit{i.e.} algorithms that run fast, but may not
return the best location\cite{Waterman1984}. Heuristic mapping algorithms
thus have three possible outcomes: $i.$ the DNA fragment is mapped to the
best location, $ii.$ the DNA fragment is mapped to another location,
$iii.$ the DNA fragment is not mapped at all.

@@

If seeds can be discovered fast, we can quickly zoom in on a small set of
candidate locations and reduce the input size of the exact but slow
algorithm. The disadvantage is that the target location is not always in
the candidate set. In this case, it cannot be discovered and the outcome
is either a false positive or a false negative.

This imposes a trade-off between speed and accuracy between each candidate
location must be tested with a slow algorithm. If the seeding step is set
to produce many candidates, the target is likely to be among them and it
will be discovered. The downside is that this will take long. Conversely,
if the seeding step is set to produce few candidates, the process will
run faster but the target is less likely to be in the candidate list so it
is less likely to be discovered.

@@

This framework was developed for the best location problem, but it also
applies to the true location problem. Actually, the process is exactly the
same but the outcomes $i.$ and $ii.$ are different. The first, called a
``true positive'' or an ``on-target hit'', is that the DNA fragment is
mapped to the \emph{true} location; the second, called a ``false
positive'' or an ``off-target hit'' is that the DNA fragment is mapped
somewhere else in the genome. A true positive thus requires that the
target (the true location) is in the candidate set, and that it has the
best score. For now, we will assume that the true location is always the
best and we defer the discussion about this assumption until
section~\ref{sec:not_the_best}.

@@

In general, seeds do not have to be exact matches\cite{pmid20460430}. For
instance, the algorithm PatternHunter\cite{pmid11934743} uses ``spaced
seeds'' that tolerate mismatches\footnote{For the anecdote, the method is
covered by a patent (reference US20070088510A1).}. However, in this
document, we will always assume that a seed is an exact match with no gap.
In other words, a seed is a perfect match between a part of the read and a
genomic site.

@@

Before going further, we need to move a practical consideration out of the
way. It stands to reason that very short matches between the read and the
genome are uninformative. For instance, $99.7\%$ of all the possible
12-mers are present in the reference human genome, so an exact match of
size 12 or lower says nothing about the location of the DNA fragment.
In what follows, we will assume that subsequences of the read can only
match the original DNA sequence and/or its duplicates, but not the rest of
the genome. This will simplify the discussion and we defer the discussion
of this assumption until section~\ref{sec:short_MEMs}

@@

\section{Neglected practical considerations}

To keep the focus on the MEM seeding process, we have ignored some
practicalities that we now have to revisit in order to evaluate the
feasibility of our approach.

\subsection{Is the true location the best one?}
\label{sec:not_the_best}

So far we have assumed that the true location is also the best one, so
that it is always discovered when it is in the candidate set. But how
likely is it that the true location is not the best one, and thus that it
is virtually impossible to discover?

Obviously, it is possible that the target has an exact duplicate. When
that happens, the read cannot be mapped unambiguously, but this case is
relatively easy to detect, so the mapper can return a low confidence score
and thus avoid making an error.

From now we assume that the read has no strict duplicate. The probability
that the read of size $k$ has $m$ errors is
\begin{equation*}
{k \choose m}p^m(1-p)^{k-m}.
\end{equation*} 
A given duplicate has a better score if its Hamming distance from the read
is less than $k$. In a first order approximation, this happens if the
duplicate matches one of the $k$ errors, and also the $L-k$ correct
nucleotides. The probability of this even is
\begin{equation*}
m(1-\mu)^{k-m}\mu/3.
\end{equation*}

There are $N$ duplicates that evolve independently, so the probability
that one of them has a better score than the target is
\begin{equation*}
1-\left(1-m(1-\mu)^{k-m}\mu/3\right)^N \approx Nm(1-\mu)^{k-m}\mu/3.
\end{equation*}

Combining the first and the last expression, we find that the probability
that the target does not have the best score is approximately
\begin{equation}
\label{eq:not_the_best}
\sum_{m=0}^k {k \choose m}p^m(1-p)^{k-m} Nm\mu/3(1-\mu)^{k-m} = 
Nk p \left(1-\mu(1-p)\right)^{k-1}\mu/3.
\end{equation}

For instance, for reads of size $k=100$ with error rate $p=0.01$, where
the target has $N=5$ duplicates with divergence $\mu=0.05$ we obtain an
estimate close to $10^{-4}$. In comparison, the probability of a false
positive when using MEM seeds with a minimum size $\gamma=17$ is
approximately $4.4\cdot10^{-4}$.

The probability that the target does not have the best score is usually
small, but it is of the same order of magnitude as the probabilities of
false positives with MEM seeds. The reason is that a common source of
false positives are the cases when the target does not have the best
score. Indeed, in this case the target is much less likely to have a MEM
seed.

In short, the event that the true location is not the best one is already
incorporated in the false positive rate and we do not recommend further
updating the estimates. In the case that one is interested in this event
\textit{per se}, formula (\ref{eq:not_the_best}) gives an approximate
probability of its occurrence. When $\mu$ is unknown, we can replace it by
its worst case value $\mu = 1/qL$, where $q=1-p$. But $N$ must be known at
least approximately because the proability increases with $N$.

In summary, the case that the true location is not the best one has a
small but not negligible probability. It is partly redundant with the
false positives generated during the MEM seeding process, and therefore it
is not straightforward to adjust the estimates accordingly. We do not
adjust them at all and consider that this case is already implicitly
accounted for.

\subsection{MEM seeds at random locations}
\label{sec:short_MEMs}

To simplify the problem, we have assumed that MEM seeds can match only the
target or one of its duplicates. In practice, MEM seeds that overlap a
sequencing error often match random locations of the genome instead of the
duplicates. When this occurs, the MEM seeds are typically small because 
the match is random.

We select MEM seeds of size at least $\gamma$, and it is important for the
validity of the theory that all the MEM seeds above the threshold are
considered candidate locations. If not, the estimates of the false
positive rate are off. The hope is that spurious MEM seeds are shorter
than $\gamma$ so that they are invisible, but this is not always the case.
And if they qualify, we need to verify them.

Such random hits can be problematic for two reasons: First, the time spent
verifying them is wasted. Second, they can cause false positives.
Fortunately, both issues can be addressed so that they do not constitute a
major problem.

To avoid spending too much time on small random MEM seeds, we can
prioritize the seeds so that the best hit is likely to be discovered first
and bail out from the alignments as soon as it appears that their score
will be worse than the best hit so far. We can even put an upper bound on
the quality of candidate so that the alignment can be skipped altogether.
For instance, if we already have a candidate at distance 1 from the read,
there is no need to verify candidates from MEM seeds that do not reach one
end of the read. Indeed, the hit must be at distance at least 2 because
there is a mismatch before the MEM seed and another one after it.

These considerations depend on the implementation of the mapper so we will
not further develop them. What matters is that random hits do not have to
impose a significant burden on the time needed to verify the candidates.

The second issue is that random hits can generate false positives. Such
false positives happen when the read does not belong to the reference
genome, or when it contains so many errors that no proper MEM seed was
found. Since the hit is not homologous to the read, the alignment score is
essentially random. The upshot is that it is easy to detect.

Indeed, if the candidate location is random, mismatches occur with
probability 3/4. If it is the right location, mismatches occur with
probability $p$. We discard the MEM seed, which is an automatic match of
size at least $\gamma$. Say that there remains $L$ nucleotides and that
$m$ of them are mismatches for the candidate location. Using Bayes
formula, we find that the probability that the location is random given
the observed alignment score is
\begin{equation}
\label{eq:bayes}
\frac{1}{1 + 4^L(p/3)^mq^{L-m}(1-\beta)/\beta},
\end{equation}
where $q=1-p$ and $\beta$ is the prior probability that the hit is
random.

One may have information about the level of contamination, allowing to
choose a meaningful value for $\beta$. For instance, sequencing runs are
often spiked by foreign DNA such as the phiX genome or other. If these
reads are not filtered automatically, $\beta$ can be set to their expected
frequency. For sequences that belong to the reference genome, random hits
are rather rare because the read must contain so many error that neither
the target nor its duplicates were discovered. The value of $\beta$ is
much smaller in this case, but nevertheless difficult to estimate.

That being said, the value of $\beta$ has little importance in practice.
Say that a read generates a hit in the genome such that 33 nucleotides
need to be aligned (the alignment status of the others was already known
from the seeding step). If the hit is random there is a 99.99\% chance
that at least 15 are mismatches. If $p = 0.01$, the denominator of
expression (\ref{eq:bayes}) is approximately $1 +
4.3\cdot10^{-18}(1-\beta)/\beta$. So unless $\beta < 10^{-17}$, the
support for the hypothesis that the hit is random is overwhelming.

Conversely, if the hit is not random, there is a 99.99\% chance that 4 or
less are mismathes. In this case, the denominator is approximately
$1+6.8\cdot10^9(1-\beta)/\beta$, so unless $\beta > 1-10^9$, the support
for the hypothesis that the hit is not random is overwhelming.

In general, one does not need to worry about the value of $\beta$ because
random hits are extremely visible when $p$ is small. For instance, one can
choose $\beta=1/2$ so that the term $(1-/\beta)/\beta$ disappears to
simplify expression (\ref{eq:bayes}). We compute this expression for every
read and add the result to the estimated probability of false positive for
the read. Either the term is negligible and it does not affect the
estimated probability, or it is strongly dominant and yields a very strong
probability that the location is incorrect, regardless of the estimate
from the MEM seeding process.

In summary, MEM seeds in random sequences happen relatively frequently,
but it is possible to minimize their computational burden. Also, they
are no cause for concern regarding false positives because they can easily
be detected using Bayes formula.

@@

The last assumption is that MEM seeds can match only the target or one of
its duplicates. This is not the case but we have shown how to gracefully
deal with this in section~\ref{sec:short_MEMs}.

The Monte Carlo approach of algorithm~\ref{alg:mcmc} is relatively
straightforward, so one may wonder why the approach with weighted
generating functions is necessary at all. The only reason is precision. To
estimate the frequency of an event by Monte Carlo sampling, it must occur
at least a few times in the simulation. For instance, with 10 million
rounds of sampling, frequencies around $1/1,000,000$ or lower cannot be
measured accurately. When not interested in rare events, it is thus a
perfectly reasonable strategy to use only Monte Carlo sampling.

For $N < 20$, the probability that a read does not contain a good seed is
the smallest, so we need a method that is accurate in this range.
Fortunately, the transfer matrix method is fast in this regime because the
matrix $M(z)$ has few rows/columns.

The proposed methods are compliant with the demand for speed of the
mapping algorithms. One needs to compute the probabilities only once for a
given value of $N$ and $\mu$ (the error rate $p$ is known and constant).
For $N > 20$, the iterative method is usually prohibitive and we need to
use Monte Carlo sampling instead. The running time depends on the desired
number of iterations, on $p$ and on the size of the reads $k$. Since those
are constant throughout the sequencing run, the method always takes the
same time (around 1 second for reads of size $100$-$200$ nucleotides on
modern hardware). The values of $N$ and $\mu$ can be binned in intervals
of interest so that there are only around $100$ pairs for a total cost of
a few minutes per run. Considering that mappers can process at most
$10,000$ reads per second per core, the time of mapping a sequencing run
of 250 million reads is over 7 hours, two orders of magnitude larger than
the time required to estimate the probabilities of error.
